---
title: 'Assignment 3'
author: 'Anthony Duong'
output:
  html_notebook:
    theme: spacelab
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

------------------------------------------------------------------------

## Assignment 4: Use of models in biology

**Part I: Population dynamics**

Recall the model of density-independent population dynamics from the
Week 11 Practical. For an organism with a current population size of n,
the population size in the next year will be:

n\^'=n(1+r)

where r is the intrinsic growth rate of the population. Recall that r is
a function of the birth and death parameters of the population (r = b
-d - bd, where b represents the average number of offspring each
individual produces during the breeding phase of the life cycle, and d
represents the proportion of individuals that die during the mortality
phase).

**Question 1.** Consider a population of perennial plants that breed in
the early spring and suffer high drought-related mortality late in the
summer. Field monitoring experiments suggest that drought leads to a 50%
decline in the population during the late summer (d = 0.5). Given this
degree of mortality, use the model to calculate how many offspring each
individual would, on average, have to produce during the breeding season
to prevent the population from declining over time. In other words,
calculate the minimum value of b that would be compatible with
population growth. [1 point]

Scoring: Full credit for providing the correct answer and showing how
the answer was obtained (i.e., show your work).

    For population to not decline over time, r ≥ 0. Thus: 
    Since; r = b - d -bd & r ≥ 0

    Therefore; 
    0 ≤ b - d - bd 
    0 ≤ b - 0.5 - 0.5b 
    0.5 ≤ b - 0.5b 
    1≤ 2b - b 
    b≥1

    Therefore, each individual would have to produce 1 offspring each year
    to stabilise the population and produce more than 1 to increase the
    population.

Suppose that you are monitoring island endemic cricket population that
has recently become threatened due to an invasive parasitoid wasp
species that is attacking its members. From observations of birth and
death rates, you estimate that the intrinsic growth rate of the cricket
population to be r ̂=-0.05, which has a 95% confidence interval of:

95% C.I. for r = [-0.01, -0.1]

Since the entire confidence interval for your estimate of r is negative,
your data imply that the population will decline over time.

**Question 2.** The current size of the cricket population is 5,000
individuals. Assume that the current conditions of parasitism do not
change and, thus, r remains constant over time. Use your point estimate
for r (r ̂=-0.05) and the model presented above to predict the amount of
time it will take for the population size to decline below 50
individuals, which is the threshold for a "critically endangered"
classification. Show how you arrived at your conclusion. [1 point]

Scoring: Full credit for providing the correct answer and showing how
the answer was obtained (i.e., submit the R code and relevant simulation
output to justify the answer OR present a detailed mathematical solution
to the question).

```{r}
r = -0.05
n.0 = 5000
n = n.0
data = n
years = 100

for(i in 1:years){
  n = n*(1 + r)
  data = c(data, n)
}

times = 0:10*(years/10)
par(mfrow = c(1, 1))
plot(0:years, data, type = "l", lwd = 3, xlab = "time (years)", ylab = "n")
points(times, n.0*(1 + r)^times, pch = 16, cex = 1, col = "red")
```

    50 = 5000*(1 - 0.05)^t
    t = 89.7

    Therefore, it'll take 90 years for the species to be classified as critically endangered

**Question 3.** There is uncertainty around your estimate of r. Suppose
that the true value of r is within the 95% confidence interval presented
above. Use the model to calculate a best-case scenario and worst-case
scenario for the amount of time it will take for the cricket population
to become critically endangered. Show how you arrived at your
conclusions. [1 point]

Scoring: Full credit for providing the correct answer and showing how
the answer was obtained (i.e., submit the R code and relevant simulation
output to justify the answer OR present a detailed mathematical solution
to the question).

```{r}
#95% C.I. for r = [-0.01, -0.1]

#worst case, lowest value of r

r = -0.1
n.0 = 5000
n = n.0
data = n
years = 50

for(i in 1:years){
  n = n*(1 + r)
  data = c(data, n)
}

times = 0:10*(years/10)
par(mfrow = c(1, 1))
plot(0:years, data, type = "l", lwd = 3, xlab = "time (years)", ylab = "n", main = 'worst case')
points(times, n.0*(1 + r)^times, pch = 16, cex = 1, col = "red")

#best case, highest value of r
r = -0.01
n.0 = 5000
n = n.0
data = n
years = 500

for(i in 1:years){
  n = n*(1 + r)
  data = c(data, n)
}

times = 0:10*(years/10)
par(mfrow = c(1, 1))
plot(0:years, data, type = "l", lwd = 3, xlab = "time (years)", ylab = "n", main = 'best case')
points(times, n.0*(1 + r)^times, pch = 16, cex = 1, col = "blue")
```

    In the best case secnario, where the growth rate r is the highest, it'll take 459 years to reach the critically endagered stage. While the worst case scenario where the growth rate r is the lowest, it'll only take 44 years.

**Part II: Evolutionary dynamics**

Recall the haploid model of evolution by natural selection from the Week
12 Prac. For a gene with two alleles (an A allele with frequency p and
fitness wA, and an a allele with frequency q = 1 -- p and fitness wa),
the allele frequency in the next generation will be:

p\^'=(pw_A)/(pw_A+(1-p) w_a )

**Question 4.** Imagine that you work in a lab that has recently carried
out an evolution experiment in which there is a rapid increase in the
frequency of an allele that they believe is beneficial in their study
system (the study system is a microbial population of haploid
individuals). Over a span of 25 generations, the allele has increased
from an initial frequency of 0.05 to a final frequency of 0.95. Members
of the lab wish to estimate the fitness advantage of individuals with
the beneficial allele relative to the fitness of individuals without the
allele (i.e., they wish to estimate wA/wa, where wA is the fitness of
individuals with the beneficial allele and wa is the fitness of
individuals with the alternative allele). Use the allele frequency data
and the haploid model to estimate the fitness advantage of individuals
with the beneficial allele. [1 point]

Scoring: Full credit for providing the correct answer and showing how
the answer was obtained (e.g., submit the R code and relevant simulation
output to justify the answer OR present a detailed mathematical solution
to the question).


**Question 5.** Now consider a population in which the fitness advantage
of the beneficial allele is 50% (i.e., wA/wa = 1.5). Suppose the
beneficial allele starts out as a single copy within a population of
100,000 individuals (its initially frequency would, therefore, be 10-5).
Assuming that the advantage of the allele remains constant over time,
how many generations will it take for it to reach a frequency of 0.5?
How long would it take to reach a frequency of 0.99? [1 point]

Scoring: Full credit for providing the correct answer and showing how
the answer was obtained (e.g., submit the R code and relevant simulation
output to justify the answer OR present a detailed mathematical solution
to the question).
```{r}
#to reach 0.5 frequency
s = -0.05/1.5
p = 0.9999
data = p

while(p > 0.5){
  p = p*(1 + s)/(1 + p*s)
  data = c(data, p)
}

par(mfrow = c(1, 1))
plot(data, type = "l", lwd = 3, xlab = "time (generations)", ylab = "p", main = '0.5 frequency')

length(data)-1

#to reach 0.99 frequency
s = -0.05/1.5
p = 0.9999
data = p

while(p > 0.01){
  p = p*(1 + s)/(1 + p*s)
  data = c(data, p)
}

par(mfrow = c(1, 1))
plot(data, type = "l", lwd = 3, xlab = "time (generations)", ylab = "p", main = '0.99 frequency')

length(data)-1
```

